Computing Vowel Elision in Yoruba Collocations

### 2023-04-14 paper.tex
Discuss Dep

\begin{figure}[h]
\caption{Our implementation of \onset}
\label{fig:impl-onset}
\begin{minted}{rust}
impl Constraint for Onset {
    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
        let syllabi = surface
            .form
            .iter()
            .filter(|seg| seg.syllable_index == SyllableIndex::Nucleus)
            .count();

@@

We now

TODO: Figure out how best to implement the algorithm described above.

TODO: Show code examples of the critical parts, one for each of the
subsections above!

>>>>

\begin{figure}[h!]
\caption{Our implementation of \onset}
\label{fig:impl-onset}
\begin{minted}[linenos]{rust}
impl Constraint for Onset {
    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
        let syllabi = surface
            .form
            .iter()
            .filter(|seg| seg.syllable_index == SyllableIndex::Nucleus)
            .count();

@@

The other significant portion of our implementation, beyond generating
candidates, is evaluating them. This, of course, is done with constraints.
Figure \ref{fig:impl-onset} shows our implementation of the \onset{}
markedness constraint. On lines 2-7, we count the number of syllable
nuclei in the input (a \texttt{SyllabifiedCandidate} from Figure
\ref{fig:impl-deletion}). On lines 9-13, we count the number of onsets.
Onsets and nuclei are calculated, along with other linguistic data, upon
the creation of a \texttt{SyllabifiedCandidate}; we merely analyze them
here. The number of violations of \onset{} is given by the number of
onsets minus the number of nuclei. This will never be a negative number,
because the algorithm that marks onsets and nuclei will only create one
onset per nucleus, at most (because this is the maximum allowed by
Yoruba).

\begin{figure}[h]
\caption{The implementation of \dep}
\label{fig:impl-dep}
\begin{minted}[linenos]{rust}
struct Dep(SyllabifiedCandidate);

impl Constraint for Dep {
    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
        let self_str: String = self.0.into();
        let surface_str: String = surface.into();

        let diff = TextDiff::from_graphemes::<String>(&self_str, &surface_str);

        diff.ops()
            .iter()
            .filter(|op| matches!(op, DiffOp::Insert { .. }))
            .count()
    }
}
\end{minted}
\end{figure}

Of course, no discussion of constraints would be complete without
\textit{faithfulness} constraints. These present an interesting challenge,
but we first provide an overview of the implementation, which is provided
in Figure \ref{fig:impl-dep}. On lines 5 and 6, we respectively turn the
original underlying form and the candidate surface form into their string
respresentations. We then use a diff algorithm on these strings on line 8,
which returns a sequence of operations which could be performed on the
original underlying form to morph it into the candidate surface form given
as input. On lines 10-13, we count the number of insertion operations
returned by the diff algorithm.

A difficult and subtle complication arises here: how does the diff
algorithm \textit{know} what counts as an insertion, deletion, or change?
For instance, changing \textipa{/ow\'o-k\'i-ow\'o/} into
\textipa{[ow\'ok\'ow\'o]} is a relatively straightforward case of
deletion. But what about \textipa{[ow\'ak\'iow\'o]}? In this case, the
diff algorithm could decide that \textipa{[\'o]} was deleted and
\textipa{[\'a]} inserted. Or it could instead say that \textipa{[\'o]}
changed into \textipa{[\'a]}. Which of these occurs is linguistically
relevant. This has not yet caused problems on real corpus data, but
when it does, a potential solution is fine-tuning the diff algorithm
to choose the more linguistically sound operations as outputs, making
it aware of autosegmental phonology when it, for example, is choosing
between a replacement operation or a insertion + deletion operations.

To this point, our descriptions have been fairly abstract, so it is best
to illustrate with an example. Given \textipa{/ow\'o-k\'i-ow\'o/} as
input, how would we generate a random candidate output form from it?
And how would we evaluate it against \onset{} and \dep?
